---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# whalesighting

<!-- badges: start -->
<!-- badges: end -->

*The Whale Hotline API* is created and maintained by [The Whale Museum](https://whalemuseum.org). Based in Washington, the museum is dedicated to promote stewardship of whales and the Salish Sea ecosystem through education and research. One of its daily functions is to run __The Whale Hotline__. Over the past 30 years, they've been collecting whales sighting data around the world. Through [The Whale Hotline API](http://hotline.whalemuseum.org/api), users can access numerous interesting whales data, and researchers can extract valuable information and observe crucial trends. 

The purpose of this project is to build a API client in the form of a R package. By installing this `whalesighting` package, users can access this database in R environment and get different kinds of whale-related information utilizing a variety of functions.  

## Installation

You can install the released version of whalesighting from github with:

```{r}
# devtools::install_github('Maxixixi/whalesighting')
library(whalesighting)
```

## Example

Here are several examples which show you how to interact with this package's 5 functions

#### Function 1: `whale_info()`
```{r example}
whale_info(10)
```

#### Function 2: `whaletype_freq()`
```{r}
whaletype_freq(50)
```

#### Function 3: `whale_des()`
```{r}
whale_des(50)
```

#### Function 4: `checkWhale()`
```{r}
checkWhale('orca')
```

#### Function 5: `whale_dist()`
```{r}
whale_dist(8.5159, -123.1524, 30.)
```








